For submission to the Journal of Integrated Pest Management

### 2021-06-08 ofv_fife_2021.Rmd
Authorities mixed up on B. confusus and B. obovatus in main text. Fixed minor grammatical errors as suggested by Word.

Chlorotic ringspots and flecking was seen affecting Liriopogons (*Liriope* and *Ophiopogon* spp.) in Leon County, FL. Nearby *A. elatior* also appeared chlorotic. Local diagnostics returned negative for common plant pathogens, therefore new samples were sent to the Florida Department of Agriculture and Consumer Services (FDACS) and USDA-ARS for identification.

@@

OFV-infected plants exhibit various symptoms depending on the infected plant species as well as the strain of the OFV associated with the infection [@Kubo2009b], but symptoms typically appear as chlorotic flecks, which ultimately coalesce into larger spots or ringspot patterns <!--(`r citef("fig_1")`, `r citef("fig_2")`)-->.

@@

As initial diagnostics were inconclusive, new samples were collected during July and August of 2020 to collect more of these putatively-infected plants with ringspot symptoms. The plants collected included *Liriope* spp. and *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea). *A. elatior* was suspected to be infected, due to both its proximity to infected *Liriope* and the presence of unusually chlorotic leaves (`r citef("fig_2")`). Upon collection, the new samples were sent to the Florida Department of Agriculture and Consumer Services (FDACS) for identification.

@@

After the initial identification by FDACS of OFV-Orc strains, mite samples were collected from symptomatic Asparagaceae in Leon county. The majority of mites collected were Tenuipalpid mites (flat mites or false spider mites), a known pest of ornamental plants, some of which are known to act as vectors for plant viruses [@Childers2003; @Childers2011].

@@

The flat mites collected were initially suspected to belong to *B. californicus* after inspection with phase contrast microscopy. Subsequent observation via DIC microscopy at FDACS agreed with this tentative identification. Unfortunately, the *B. californicus* s.l. species group, sensu @Baker1987 is suspected to contain cryptic species [@Childers2011; @Rodrigues2013]. With this in mind, new mite samples were collected from symptomatic liriopogons and *A. elatior* in Leon County and sent to USDA-ARS's Electron and Confocal Microscopy Unit for analysis. Three mite species were recovered and examined under cryo-scanning electron microscopy (Cryo-SEM): *B. californicus* s.l. (`r citef("fig_3")`), *B. obovatus* Baker and *B. confusus* Donnadieu (`r citef("fig_4")`).

@@

OFV consists of two orchid strains (OFV-Orc1 and OFV-Orc2) and two citrus strains (OFV-Cit1 and OFV-Cit2) [@Roy2020; @Beltran-Beltran2020]. The OFV strains detected in Florida are identical in gene order, content, and genome sequence to the orchid strains of OFV infecting citrus in Hawaii, Mexico, Colombia, and South Africa [@Roy2020; @Beltran-Beltran2020]. Both OFV-Orc1 and OFV-Orc2 infect citrus [@Roy2020], but none of the citrus strains have been reported from any orchid species. The *Brevipalpus* mites collected from liriopogons and *A. elatior* in Leon county were abundant on OFV-infected plants very near to citrus trees, some plants even surrounding the trunk. *B. californicus* s. l. have been reported as a pest of citrus [@Childers2003] and are often collected from citrus rinds [@Baker1949; @Baker1987], so the proximity of these mite vectors to citrus raises the question: why these trees are not currently infected with OFV-Orc? Each possibility suggest further research, but there are a few possible explanations. It is important to note the uncertainty surrounding the vector for OFV-Orc. There are three mite species which have been recovered from OFV-Orc infected plants: *B. californicus* (the most likely culprit), *B. obovatus*, and *B. confusus*. Each species has its own unique biology and all have been implicated with a variety of different hosts. This suggests that the spread of OFV-Orc would be a function of various combinations of a number of potential factors, possibly including: host preferences, vectorial capacity, viral propagation/circulation in the vector, viral acquisition times, and feeding times required for transmission. Another possiblity is that the *B. californicus* which we find on liriopogons and *A. elatior* are not actually the same species as those found on citrus, and instead represent 2-3 different cryptic species.

Detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivated *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and almost eradicated by the mid-1960s [@Childers2003; @Knorr1968a; @Knorr1968b]. An examination of herbarium specimens of Florida citrus found that this historical virus, Citrus leprosis dichorhavirus-N0, is distantly related to the modern strains of OFV [@Hartung2015; @Kitajima2011a; @Roy2020]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) associated with leprosis-like symptoms highlights the threat of different strains of OFV on citrus; which will be a definite concern to the US multi-billion dollar citrus industry. *B. californicus*, as well as *B. yothersi* (Baker), are both known vectors of dichorhaviruses (OFV) [@Kondo2003; @Beltran-Beltran2020] and *B. obovatus* is a suspected vector as well [@Childers2003]. All three mite species/complexes are present in Florida [@Childers2003; @Akyazi2017] (`r citef("fig_4")`), and are difficult to identify by non-experts, or without advanced methodologies. DNA barcoding [@Armstrong2005] or a similarly simple and accurate method for identification of these mite complexes is vital to determine which of these species are responsible for transmission of OFV-Orc, and therefore which mite populations need to be monitored or controlled. By doing so, we can determine the risk OFV-Orc represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.

>>>>

Chlorotic ringspots and flecking were seen affecting Liriopogons (*Liriope* and *Ophiopogon* spp.) in Leon County, FL. Nearby *A. elatior* also appeared chlorotic. Local diagnostics returned negative for common plant pathogens, therefore new samples were sent to the Florida Department of Agriculture and Consumer Services (FDACS) and USDA-ARS for identification.

@@

OFV-infected plants exhibit various symptoms depending on the infected plant species as well as the strain of the OFV associated with the infection [@Kubo2009b], but symptoms typically appear as chlorotic flecks, which ultimately coalesce into ringspot patterns.

@@

As initial diagnostics were inconclusive, new samples were collected during July and August of 2020 to collect more of these putatively infected plants with ringspot symptoms. The plants collected included *Liriope* spp. and *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea). *A. elatior* was suspected to be infected, due to both its proximity to infected *Liriope* and the presence of unusually chlorotic leaves (`r citef("fig_2")`). Upon collection, the new samples were sent to the Florida Department of Agriculture and Consumer Services (FDACS) for identification.

@@

After the initial identification by FDACS of OFV-Orc strains, mite samples were collected from symptomatic Asparagaceae in Leon County. Most mites collected were Tenuipalpid mites (flat mites or false spider mites), a known pest of ornamental plants, some of which are known to act as vectors for plant viruses [@Childers2003; @Childers2011].

@@

The flat mites collected were initially suspected to belong to *B. californicus* after inspection with phase contrast microscopy. Subsequent observation via DIC microscopy at FDACS agreed with this tentative identification. Unfortunately, the *B. californicus* s.l. species group, sensu @Baker1987 is suspected to contain cryptic species [@Childers2011; @Rodrigues2013]. New mite samples were collected from symptomatic liriopogons and *A. elatior* in Leon County and sent to USDA-ARS's Electron and Confocal Microscopy Unit for analysis. Three mite species were recovered and examined under cryo-scanning electron microscopy (Cryo-SEM): *B. californicus* s.l. (`r citef("fig_3")`), *B. obovatus* Donnadieu and *B. confusus* Baker (`r citef("fig_4")`).

@@

OFV consists of two orchid strains (OFV-Orc1 and OFV-Orc2) and two citrus strains (OFV-Cit1 and OFV-Cit2) [@Roy2020; @Beltran-Beltran2020]. The OFV strains detected in Florida are identical in gene order, content, and genome sequence to the orchid strains of OFV infecting citrus in Hawaii, Mexico, Colombia, and South Africa [@Roy2020; @Beltran-Beltran2020]. Both OFV-Orc1 and OFV-Orc2 infect citrus [@Roy2020], but none of the citrus strains have been reported from any orchid species. The *Brevipalpus* mites collected from liriopogons and *A. elatior* in Leon County were abundant on OFV-infected plants very near to citrus trees, some plants even surrounding the trunk. *B. californicus* s. l. has been reported as a pest of citrus [@Childers2003] and are often collected from citrus rinds [@Baker1949; @Baker1987], so the proximity of these mite vectors to citrus raises the question: why these trees are not currently infected with OFV-Orc? Each possibility suggests further research, but there are a few possible explanations. It is important to note the uncertainty surrounding the vector for OFV-Orc. There are three mite species which have been recovered from OFV-Orc infected plants: *B. californicus* (the most likely culprit), *B. obovatus*, and *B. confusus*. Each species has its own unique biology, and all have been implicated with a variety of different hosts. This suggests that the spread of OFV-Orc would be a function of various combinations of several potential factors, possibly including host preferences, vectorial capacity, viral propagation/circulation in the vector, viral acquisition times, and feeding times required for transmission. Another possibility is that the *B. californicus* which we find on liriopogons and *A. elatior* are not actually the same species as those found on citrus, and instead represent 2-3 different cryptic species.

Detecting OFV in Florida represents a concern for horticulturists who grow orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae species which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including cultivated *Vanilla* in southern Florida [@Chambers2019] and the famous Ghost Orchid, [*Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe]. Citrus leprosis was present in Florida during the 1860’s and almost eradicated by the mid-1960s [@Childers2003; @Knorr1968a; @Knorr1968b]. An examination of herbarium specimens of Florida citrus found that this historical virus, Citrus leprosis dichorhavirus-N0, is distantly related to the modern strains of OFV [@Hartung2015; @Kitajima2011a; @Roy2020]. The recent detection of OFV-Orc1 in South Africa [@Cook2019] in *C. sinensis* (Navel and Valencia orange) and OFV-Orc2 in Hawaii [@Velarde2021] in *C. reticulata* (mandarin) and *C. jambhiri* (rough lemon) associated with leprosis-like symptoms highlights the threat of different strains of OFV on citrus, which will be a definite concern to the US multi-billion-dollar citrus industry. *B. californicus*, as well as *B. yothersi* (Baker), are both known vectors of dichorhaviruses (OFV) [@Kondo2003; @Beltran-Beltran2020] and *B. obovatus* is a suspected vector as well [@Childers2003]. All three mite species/complexes are present in Florida [@Childers2003; @Akyazi2017] (`r citef("fig_4")`), and are difficult to identify by non-experts, or without advanced methodologies. DNA barcoding [@Armstrong2005] or a similarly simple and accurate method for identification of these mite complexes is vital to determine which of these species are responsible for transmission of OFV-Orc, and therefore which mite populations need to be monitored or controlled. By doing so, we can determine the risk OFV-Orc represents for the native plants, agriculture and the ornamental/landscaping industries of Florida and the surrounding regions.
